An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2017-03-30 uai2017/paper.tex
fix organization

We note that this approach is fundamentally different from prior
work. It makes no assumptions and requires no global bounds on the
posterior parameter distribution in order to work. It is exact in the
limit as batch size increases by the Central Limit Theorem. This is
not true of ~\citep{cutting_mh_2014} and~\citep{icml2014c1_bardenet14}
which use tail bounds and provide only approximate tests even with
arbitratily large batches of data. Our test is also exact under the
assumptions of \citep{cutting_mh_2014} that per-sample parameter
posteriors are normally distributed about the population mean.  Rather
than tail bounds, our approach uses moment estimates from the data to
determine how far the minibatch posteriors deviate from a normal
distribution.  These bounds carry through to the overall accuracy of
the test.

@@

  \item Utilize Hamiltonian Dynamics for proposals and tests. Here the
    dynamics itself provide shaping to the posterior distribution, and
    the M-H is only needed to correct for quantization error. In terms
    of the information carried by the samples, this approach is not
    limited by the data in a particular minibatch since momentum is
    carried over time and ``remembered'' across multiple
    minibatches.
\end{enumerate}

@@

To address this challenge,~\citep{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate Metropolis-Hasting tests using sequential hypothesis testing. Each
iteration, they start with a small minibatch and test whether $\theta'$ should
be accepted based on approximating $u<\alpha(\theta,\theta')$. If the
approximate test cannot decide with sufficient confidence, the minibatch grows
and the test repeats. This process continues until a decision. The bounds depend
on either an asymptotic Central Limit Theorem~\citep{cutting_mh_2014} or a
concentration bound~\citep{icml2014c1_bardenet14}. We refer to these algorithms,
respectively, as {\sc AustereMH} and {\sc MHSubLhd}. While both show useful
reductions in the number of samples required, they suffer from two drawbacks:
having to compute all log likelihood tokens each iteration to compute certain
statistics,\footnote{Thus, these are not technically minibatch methods.} (see
Section~\ref{sec:analysis}), and resolving small log likelihood ratio
differences between the minibatch and full batch versions. We discuss a
worst-case scenario in Section~\ref{ssec:gaussian_example}.

@@

% Daniel: use this algorithm style instead
\begin{algorithm}[tb]
\caption{Our acceptance test, {\sc MHminibatch}.}
\label{alg:our_algorithm}
\begin{algorithmic}
    \STATE {\bfseries Input:} number of samples $T$, minibatch size $m$, error
        bound $\delta$, pre-computed correction $C_1(X)$ distribution, initial
        sample $\theta_1$.
    \STATE {\bfseries Output:} a chain of $T$ samples
        $\{\theta_1,\ldots,\theta_T\}$.
    \FOR{$t=1$ {\bfseries to} $T$}
        \STATE -Propose a candidate $\theta'$ from proposal
            $q(\theta' | \theta_t)$.
        \STATE -Draw a minibatch of $m$ points $\{x_1^*,\ldots,x_m^*\}$.
        \STATE -Compute $\Delta^*(\theta_t,\theta')$ and sample variance
            $s^2_{\Delta^*}$.
        \STATE -Estimate moments $\mE[|\Lambda_i-\Lambda|]$ and
            $\mE[|\Lambda_i-\Lambda|^3]$ from the sample, and error $\epsilon$
            from Corollary~\ref{cor:our_bound_delta_prime}.
        \WHILE{$s^2_{\Delta^*} \ge 1$ {\bf or} $\epsilon > \delta$}
            \STATE -Draw $m$ more samples to augment the minibatch, update
                $\Delta^*$, $s^2_{\Delta^*}$ and $\epsilon$ estimates.
        \ENDWHILE
        \STATE -Draw $X_{\rm nc} \sim \mathcal{N}(0,1-s^2_{\Delta^*})$ and
            $X_{\rm corr} \sim C_1(X)$.
    \IF{$\Delta^* + X_{\rm nc} + X_{\rm corr} > 0$}
        \STATE -Accept the candidate, $\theta_{t+1} = \theta'$.
    \ELSE
        \STATE -Reject and re-use the old sample, $\theta_{t+1} = \theta_t$.
    \ENDIF
    \ENDFOR
\end{algorithmic}
\end{algorithm}

@@

We now derive error bounds for our M-H test and the target distribution it
generates. In the most similar prior works,~\citep{cutting_mh_2014}, CLT
asymptotic arguments are used to show that their approximate acceptance test
error (with respect to full-batch M-H) tends to zero as batch size increases,
but no quantitative bounds are given. In addition, to design a test that
minimizes data usage given an error bound, they rely on computing a standardized
mean $\mu_{\rm std}$ each iteration. As the value of $\mu_{\rm std}$ depends on
all $\log p(x_i|\theta)$ and $\log p(x_i| \theta')$ terms, the test is
technically not a minibatch test; \citep{cutting_mh_2014} propose a conservative
variant which assumes $\mu_{\rm std}=0$ and avoids the need to evaluate all
likelihood terms. We refer to this as {\sc AustereMH(c)} and the default,
non-conservative variant as {\sc AustereMH(nc)}.

For {\sc MHSubLhd}, the test requires direct bounds on the log likelihood ratio,
which for general distributions requires knowing $p(x_i | \theta)$ and $p(x_i |
\theta')$ for all $N$ samples. In~\citep{icml2014c1_bardenet14}, explicit bounds
are given and depend on
\begin{equation}\label{eq:bad_bound}
    C_{\theta, \theta'} = \max_{1\leq i\leq N}|\log p(x_i | \theta') - \log p(x_i | \theta)|.
\end{equation}
It is possible to analytically determine $C_{\theta,\theta'}$ for logistic
regression and other simple models, but it is difficult to do so for more
complicated models, so one generally needs to use all $p(x_i | \theta')$
terms\footnote{The sample code provided by~\citep{icml2014c1_bardenet14}
computes $C_{\theta,\theta'}$ by traversing the entire data, thus providing no
performance advantage over the complete test.} to get $C_{\theta,\theta'}$, like
the {\sc AustereMH(nc)} test.  In contrast, we use quantitative forms of the CLT
which rely on measurable statistics from a \emph{single} minibatch.

In Section~\ref{ssec:delta_star_distribution}, we present bounds on the absolute
and relative error (in terms of the CDFs) of the distribution of $\Delta^*$
versus a Gaussian. We then show in Section~\ref{ssec:preserve_bounds} that these
bounds are preserved after the addition of other random variables (e.g., $X_{\rm
nc}$ and $X_{\rm corr}$). It then follows that the acceptance test has the same
error bound.

@@

\section{EXPERIMENTS}\label{sec:experiments}

>>>>

We note that this approach is fundamentally different from prior
work. It makes no assumptions about the form of, and requires no global bounds on the
posterior parameter distribution. It is exact in the
limit as batch size increases by the Central Limit Theorem. This is
not true of ~\citep{cutting_mh_2014} and~\citep{icml2014c1_bardenet14}
which use tail bounds and provide only approximate tests even with
arbitratily large batches of data. Our test is also exact under the
assumptions of \citep{cutting_mh_2014} that the log probability ratios
of batches are normally distributed about their mean.  Rather
than tail bounds, our approach uses moment estimates from the data to
determine how far the minibatch posteriors deviate from a normal
distribution.  These bounds carry through to the overall accuracy of
the test.

@@

  \item Utilize Hamiltonian Dynamics for proposals and tests. Here the
    dynamics itself provide shaping to the posterior distribution, and
    the M-H test is only needed to correct quantization error. In terms
    of the information carried by the samples, this approach is not
    limited by the data in a particular minibatch since momentum is
    carried over time and ``remembered'' across multiple
    minibatches.
\end{enumerate}

@@

To address this
challenge,~\citep{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate Metropolis-Hasting tests using sequential hypothesis
testing. At each iteration, a subset of data is sampled and used to
test whether to accept $\theta'$ using an approximation to
$\alpha(\theta,\theta')$. If the approximate test does not yield a
decision, the minibatch size is increased and the test repeated. This
process continues until a decision. These methods either invoke the
asymptotic CLT and assume that finite  batch errors are normally
distributed~\citep{cutting_mh_2014} or use a concentration
bound~\citep{icml2014c1_bardenet14}. We refer to these algorithms,
respectively, as {\sc AustereMH} and {\sc MHSubLhd}. While both show
useful reductions in the number of samples required, they suffer from
two drawbacks: (i) They are approximate, and always yield a decision
with a finite error, (ii) They both require exact, dataset-wide bounds
that depend on $\theta$.
\footnote{We obtained the authors code for both and found that they scanned the entire
dataset at each step to obtain these estimates.} (see
Section~\ref{sec:analysis}). We discuss a
worst-case scenario in Section~\ref{ssec:gaussian_example}.

@@

% Daniel: use this algorithm style instead
\begin{algorithm}[tb]
\caption{{\sc MHminibatch} acceptance test.}
\label{alg:our_algorithm}
\begin{algorithmic}
    \STATE {\bfseries Input:} number of samples $T$, minibatch size $m$, error
        bound $\delta$, pre-computed correction $C_1(X)$ distribution, initial
        sample $\theta_1$.
    \STATE {\bfseries Output:} a chain of $T$ samples
        $\{\theta_1,\ldots,\theta_T\}$.
    \FOR{$t=1$ {\bfseries to} $T$}
        \STATE -Propose a candidate $\theta'$ from proposal
            $q(\theta' | \theta_t)$.
        \STATE -Draw a minibatch of $m$ points $\{x_1^*,\ldots,x_m^*\}$.
        \STATE -Compute $\Delta^*(\theta_t,\theta')$ and sample variance
            $s^2_{\Delta^*}$.
        \STATE -Estimate moments $\mE[|\Lambda_i-\Lambda|]$ and
            $\mE[|\Lambda_i-\Lambda|^3]$ from the sample, and error $\epsilon$
            from Corollary~\ref{cor:our_bound_delta_prime}.
        \WHILE{$s^2_{\Delta^*} \ge 1$ {\bf or} $\epsilon > \delta$}
            \STATE -Draw $m$ more samples to augment the minibatch, update
                $\Delta^*$, $s^2_{\Delta^*}$ and $\epsilon$ estimates.
        \ENDWHILE
        \STATE -Draw $X_{\rm nc} \sim \mathcal{N}(0,1-s^2_{\Delta^*})$ and
            $X_{\rm corr} \sim C_1(X)$.
    \IF{$\Delta^* + X_{\rm nc} + X_{\rm corr} > 0$}
        \STATE -Accept the candidate, $\theta_{t+1} = \theta'$.
    \ELSE
        \STATE -Reject and re-use the old sample, $\theta_{t+1} = \theta_t$.
    \ENDIF
    \ENDFOR
\end{algorithmic}
\end{algorithm}

@@

We now derive error bounds for our M-H test and the target distribution it
generates. 
In Section~\ref{ssec:delta_star_distribution}, we present bounds on the absolute
and relative error (in terms of the CDFs) of the distribution of $\Delta^*$
versus a Gaussian. We then show in Section~\ref{ssec:preserve_bounds} that these
bounds are preserved after the addition of other random variables (e.g., $X_{\rm
nc}$ and $X_{\rm corr}$). It then follows that the acceptance test has the same
error bound.

@@

\section{EXPERIMENTS}\label{sec:experiments}
In one of the most closely-related prior
works,~\citep{cutting_mh_2014}, asymptotic CLT arguments are used to
argue that a modified standard M-H test can be used on subsets of data
with modest error. This assumes knowledge of dataset-wide mean
$\mu_{\rm std}$ each iteration (it depends on $\theta$). Determining
$\mu_{\rm std}$ exactly requires a scan over the entire dataset (so
the method in practice provides no data reduction), or some
model-specific bounds must be derived.  \citep{cutting_mh_2014} also
propose a conservative variant which assumes $\mu_{\rm std}=0$ and
avoids the scan.  We refer to the conservative version as {\sc
  AustereMH(c)} and the non-conservative variant as {\sc
  AustereMH(nc)}. We analyze both in the sequel.

In another related work~\citep{icml2014c1_bardenet14} use concentration bounds
and a similar modification to the standard M-H test ({\sc MHSubLhd} method). 
For {\sc MHSubLhd}, the test requires direct bounds on the log likelihood ratio,
which for general distributions requires knowing $p(x_i | \theta)$ and $p(x_i |
\theta')$ for all $N$ samples. In~\citep{icml2014c1_bardenet14}, explicit bounds
are given and depend on
\begin{equation}\label{eq:bad_bound}
    C_{\theta, \theta'} = \max_{1\leq i\leq N}|\log p(x_i | \theta') - \log p(x_i | \theta)|.
\end{equation}
It is possible to analytically determine $C_{\theta,\theta'}$ for logistic
regression and other simple models, but it is difficult to do so for more
complicated models, so one generally needs to use all $p(x_i | \theta')$
terms\footnote{The sample code provided by~\citep{icml2014c1_bardenet14}
computes $C_{\theta,\theta'}$ by traversing the entire data, thus providing no
performance advantage over the complete test.} to get $C_{\theta,\theta'}$, like
the {\sc AustereMH(nc)} test.  In contrast, we use quantitative forms of the CLT
which rely on measurable statistics from a \emph{single} minibatch.
